---
title: "Analysis"
---

```{r}
#| echo: false
#| message: false
#| output: false
# install necessary libraries
library(readxl)
library(tidyverse)
library(writexl)
library(stringr)
library(ggthemes)
library(plotly)
library(formattable)
library(DT)
library(ggplot2)
```

```{r}
#| echo: false
#| message: false
#| output: false
# read the excel file as a csv
top5 <- read_excel("Top5Leagues_Payroll.xlsx")
```

### General 
Across the data set, we can a positive correlation between payroll rank and standings. Lookly closely there are notable examples like Leicester City's 15/16 EPL title.

```{r}
#| echo: false
#| message: false
#| output: false
# create data frame where I can extract payroll rank vs standing 
d <- top5 |>
  select("Season", "Rk", "Squad", "Annual Wages", "Standing") |>
  mutate(annual = str_extract(`Annual Wages`, "\\$[0-9,]+")) |>
  mutate(annual = as.numeric(gsub("[$,]", "", annual))) |>
  select(-`Annual Wages`) |>
  summarise(.by = Season, Rk = Rk, Squad = Squad, Standing = Standing, annual = annual, percent = annual/sum(annual))
```

```{r}
#| echo: false
#| message: false
#| output: false
#| warning: false
# create ggplot object of payroll rank vs standing
d_1 <- d |>
  ggplot(aes(x = Rk, y = Standing)) +
  geom_jitter(color = "#30a2da", alpha = 0.5, aes(text = paste("Season:", Season, "\n", 
                                                "Payroll:", currency(annual), "\n",
                                                "Standing:", Standing, "\n",
                                                "Rank:", Rk, "\n",
                                                "Team:", Squad))) +
  scale_x_reverse(breaks = seq(20, 1, by = -1)) +
  scale_y_reverse(breaks = seq(20, 1, by = -1)) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x, color = "#fc4f30") +
  labs(title = "Payroll Rank vs. Season Standing \n in the Top Five European Leagues",
       x = "Payroll Rank",
       y = "Season Standing",
       caption = "Source: FBREF") +
  theme_fivethirtyeight() +
  theme(axis.title.x = element_text()) +
  theme(axis.title.y = element_text()) +
  theme(plot.caption = element_text())
```

```{r}
#| echo: false
#| message: false
#| warning: false
# plot interactive graph
d_1 |>
  ggplotly(tooltip = "text") |>
  layout(title = list(font = list(family = "Arial")))
```

### Where it matters
In European soccer competition, the top four and bottom three standings are important. For the top leagues, a top four place secures Champions League qualification, the most prestigious club soccer tournament; a bottom three place means relegation (demotion to the second division). 

```{r}
#| echo: false
#| message: false
#| output: false
# create and save a tibble, where the first season is 2013 and selects teams of the top four and bottom three (top four usually Champions League qualification, bottom three relegationn)
b <- top5 |>
  select(-c("# Pl", "Weekly Wages", "% Estimated", "Squad (again)")) |>
  mutate(annual = str_extract(`Annual Wages`, "\\$[0-9,]+")) |>
  mutate(annual = as.numeric(gsub("[$,]", "", annual))) |>
  select(-`Annual Wages`) |>
  filter(Season >= 2013, Standing <= 4 | Standing >= 18) |>
  mutate(top = Standing <= 4) |>
  mutate(top = ifelse(top == TRUE, "Top 4", "Bottom 3")) 
```

```{r}
#| echo: false
#| message: false
#| output: false
# create ggplot object, mapping top team and bottom teams in standing and their payrolls
c <- b |>
  ggplot(aes(x = Season, y = annual, color = top, text = paste("Season:", Season, "\n", 
                                                               "Payroll:", currency(annual), "\n",
                                                               "Standing:", Standing, "\n",
                                                               "Team:", Squad))) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = seq(2013, 2022, by = 1)) +
  labs(title = "Payroll of Top Four vs Bottom Three \n Performing Teams",
       subtitle = "The payroll gap of top from bottom has steadily increased",
       y = "Payroll",
       color = "") +
  theme_fivethirtyeight()
```

```{r}
#| echo: false
#| message: false
#| warning: false
# convert to interactive plot
c |>
  ggplotly(tooltip = "text") |>
  layout(title = list(font = list(family = "Arial")),
         legend = list(font = list(family = "Arial")))
```

### Which teams are spending wisely?
To categorize how teams are performing relative to their the payroll rank, I took the difference of standing and payroll rank each year as a measurement. A positive differece means that a team over performed that specific year. The total column represents the sum of the difference in all the years the team appeared in the data set, and rate is the average difference.

```{r}
#| echo: false
#| message: false
#| output: false
# create data frame to measure a team's performance relative to payroll rank, diff is difference between payroll rank and standing (+ is good)
differential <- top5 |>
  select(-c("# Pl", "Weekly Wages", "% Estimated", "Squad (again)")) |>
  mutate(annual = str_extract(`Annual Wages`, "\\$[0-9,]+")) |>
  mutate(annual = as.numeric(gsub("[$,]", "", annual))) |>
  select(-`Annual Wages`) |>
  mutate(diff = Rk - Standing) |>
  summarise(.by = Squad, total = sum(diff), years = n()) 
```


```{r}
#| echo: false
#| message: false
#| output: false
# add rate column since not all teams appear in the data the same amount of times
diff_1 <- differential |>
  mutate(rate = total / years) |>
  arrange(desc(rate)) 

#create interactive data frame
diff_DT <- diff_1 |>
  datatable()
```

```{r}
#| echo: false
#| message: false
# Display datatable
diff_DT |>
  formatStyle(columns = names(diff_1), fontFamily = "Arial")
```

