---
title: "Overview"
execute: 
  echoe: false
  message: false
---

### Context


```{r}
#| echo: false
#| message: false
#| output: false
# upload necessary libraries
library(readxl)
library(tidyverse)
library(writexl)
library(stringr)
library(ggthemes)
```

```{r}
# read the excel file as a csv
top5 <- read_excel("Top5Leagues_Payroll.xlsx")
```

```{r}
#| echo: false
#| message: false
#| output: false
# select the necessary columns and create a data frame where there is the total money spent of each league across seasons
a <- top5 |>
  select(-c("# Pl", "Weekly Wages", "% Estimated", "Squad (again)")) |>
  mutate(annual = str_extract(`Annual Wages`, "\\$[0-9,]+")) |>
  mutate(annual = as.numeric(gsub("[$,]", "", annual))) |>
  select(-`Annual Wages`) |>
  filter(Season >= 2013) |>
  summarise(.by = c(League, Season), total = sum(annual))
```

```{r}
# plot the graph
a |>
  ggplot(aes(x = Season, y = total, color = League)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = seq(2013, 2022, by = 1)) +
  labs(title = "Total Annual Wages for Each League",
       subtitle = "Each league has steadily increased spending",
       color = NULL) +
  theme_fivethirtyeight()
```

Soccer clubs have been rapidly spending more on their squads. The data was collected from https://fbref.com/en/; since payrolls and standings were stored in different locations and separated by years, the data was compiled into a single excel file. Data includes the top five European leagues from mostly the 2013 to 2022 seasons. I estimated a Bayesian linear model with standing as the dependent variable and payroll rank as the independent variable.

