---
title: "Overview"
execute: 
  echoe: false
  message: false
---

### Context
With the global commercialization of soccer, clubs have been significantly spending more on their squads. Takeovers from American investors and controversy concerning foreign investment has sparked topics about teams' sources of wealth. Across many sports, there have been discussions about the implementation of salary caps.

```{r}
#| echo: false
#| message: false
#| output: false
# upload necessary libraries
library(readxl)
library(tidyverse)
library(writexl)
library(stringr)
library(ggthemes)
```

```{r}
#| echo: false
#| message: false
#| output: false
# read the excel file as a csv
top5 <- read_excel("Top5Leagues_Payroll.xlsx")
```

```{r}
#| echo: false
#| message: false
#| output: false
# select the necessary columns and create a data frame where there is the total money spent of each league across seasons
a <- top5 |>
  select(-c("# Pl", "Weekly Wages", "% Estimated", "Squad (again)")) |>
  mutate(annual = str_extract(`Annual Wages`, "\\$[0-9,]+")) |>
  mutate(annual = as.numeric(gsub("[$,]", "", annual))) |>
  select(-`Annual Wages`) |>
  filter(Season >= 2013) |>
  summarise(.by = c(League, Season), total = sum(annual))
```

```{r}
#| echo: false
#| message: false
# plot the graph
a |>
  ggplot(aes(x = Season, y = total, color = League)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = seq(2013, 2022, by = 1)) +
  labs(title = "Total Annual Wages for Each League",
       subtitle = "Each league has steadily increased spending",
       caption = "Source: FBREF",
       color = NULL) +
  theme_fivethirtyeight()
```
### Summary
The data was collected from [fbref](https://fbref.com/en/); since payrolls and standings were stored in different locations and separated by years, the data was compiled into a single excel file. Data includes the top five European leagues from mainly the 2013 to 2022 seasons. I estimated a Bayesian linear model with standing as the dependent variable and payroll rank as the independent variable.

